Parallel Multi-objective optimization using co-evolutionary islands and search space overlapping

### 2016-06-15 hpmoon-hpca.tex
Added @pacastillo and Julio comments

\def\volumeyear{2015}

@@

%\author{Pablo Garc\'ia-S\'anchez, Julio Ortega, Jes\'us Gonz\'alez, \\Pedro Angel Castillo and Juan Juli\'an Merelo\affilnum{1}}
\author{ANONYMOUS FOR BLIND REVIEW \affilnum{1}}
\affiliation{\affilnum{1} ANONYMOUS FOR BLIND REVIEW}
%\affiliation{\affilnum{1}Dept. of Computer Architecture and Computer Technology, University of Granada, Spain}

@@

\corrauth{ANONYMOUS FOR BLIND REVIEW.}

@@

In our preliminary work in \citep{Garcia16hpmoon} we tried to demonstrate that using overlapped sections of the chromosome in a coevolutionary multi-objective island-based algorithm can improve the quality of the solutions in the same computing time when the number of islands increases. We discovered that the most adequate overlapping size may vary depending on a relation between the number of islands and the individual size. Moreover, the experiments were conducted in a mono-processor and synchronous island model. This motivates us to continue this research line, but this time, proposing a new method to automatically adapt the size of the overlapped sections, but also, performing all the experiments in a real cluster with up to 128 nodes, with extra number of islands.

@@

The rest of the paper is organized as follows: after a background in parallelization in MOEAs, 
the compared algorithms and the used methodology (Section 3) are presented. %PABLO: no puedo usar \ref{sec:met} porque la plantilla de esta revista no lo coge :/
Then, the results of the experiments are shown (Section 4), followed by conclusions and suggestions for future work lines.

@@

In our previous work \citep{Garcia16hpmoon} we used some of the previous ideas to compare two co-evolutionary dMOEAs. The first one divided the chromosome in $P$ sections, being $P$ the number of islands. Every island $p$ only performed the mutation and crossover in that part ($p_{th}$) of the chromosome, while the fitness was calculated using the whole individual. After a certain number of generations, individuals were migrated randomly to other islands. The performance metrics were calculated at the end of the run. The second method, coevolution with overlapping islands, used the $p_{th-1}$ and $p_{th+1}$ sections of each chromosome, besides the $p_{th}$ part. During the same amount of time both methods obtained better results than a baseline algorithm that dealt with the whole chromosome in each island for crossover and mutation. We discovered that the performance using one or another method depends on the number of sections of the individuals and the number of islands used. This motivates us to find a new adaptive method to select this number of sections of the chromosome to use, depending on the number of islands. Moreover, previous experiments were performed on a single processor synchronous island model with a limited number of islands (8, 32 and 128). In this paper, we have used 8, 16, 32, 64 and 128 islands, and this time, using a real cluster. Therefore, at the same time we are proposing an adaptive method to divide the individual search space, we are validating the previous methods.

@@

%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%  METHODOLOGY  %%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%
\section{Methodology and experimental setup}
\label{sec:met}

This section describes the quality indicators used and the methods compared. The chosen quality indicators, are:

@@

These metrics have been used extensively, especially in some of the papers presented previously in Section 2.

%% --------------------------------------------------------------

\subsection{Baseline algorithm}

This algorithm is a regular NSGA-II algorithm distributed along a number $P$ of islands. After a fixed number of generations, one individual is migrated to another random island. At the end of the run, all PFs of all islands are aggregated in a new one to compute the quality measures. This method as been chosen as it do not require synchronization of the global PF every certain number of generations, as other methods described in Section 2, and because it can {\color{red} QUE MAS SE PUEDE PONER AQUI PARA JUSTIFICARLO?}
%% some justification on why this was chosen - JJ PABLO: Added 
%
%% And some liaison to next subsection - JJ
This method will be compared to different coevolutionary approaches, that are similar to the baseline, with the exception on the decision space to explore in each island.
\subsection{Coevolutionary algorithms compared}
Three different coevolutionary methods have also been used. In all these methods, each island only performs crossover and mutation in specific sections of the individuals.

As in the baseline, every certain number of generations, an individual is migrated to another random island.  In the new island, this individual will be considered as one of the others in the island, crossed and mutated in the same way, depending of the island identifier. Note that, on the contrary of other works such as the ones described by Talbi et al. \citep{Talbi08Parallel} all the islands deal with complete chromosomes for fitness calculation, so our approach can deal with no decomposable problems.

\subsubsection{Coevolution with disjoint islands (D)} 
In this approach, each individual of size $L$ is split into $P$ chunks of size $L/P$. Every island $p$ only performs crossover and mutation on the $p_{th}$ part of the individuals. Figure \ref{fig:disjoint} describes this approach.

\subsubsection{Coevolution with overlapping islands (O)}
This approach is similar to the previous one, but every island also uses the $p+1$ and $p-1$ (module size) chunks of the individual for crossover and migration. Therefore, some kind of overlapping of the crossed and mutated parts exists between islands. Figure \ref{fig:overlapping} shows the affected parts of the individuals in
each island.

\subsubsection{Adaptive coevolution with overlapping islands (A)} 
As in the previous method, the sections to address are overlapped, but instead using one extra section on each side of the $p$ section ($p+1$ and $p-1$), it uses $c$ fragments on each side ($p+c$ and $p-c$), being $c$ a value depending on the number of islands. As a first approach to automatically calculate this value the results shown in \citep{Garcia16hpmoon} have been used as a base to obtain $c$. In that work, the overlapped method ($c=1$) obtained better results when the number of islands were $>$8. So, we have used the formula $c=round(0.2*P-1)/2$ to calculate the extra sections to overlap. Therefore, when $P=8$ then $c=0$ (equivalent to D), when $P=16$ then $c=1$ (equivalent to O), and so on. Figure \ref{fig:adaptive} explains this method.

The four approaches have been run with two different chromosome lengths ($L$): 512 and 2048. Different number of islands ($P$) have also been compared: 8, 16, 32, 64 and 128. This maximum number of islands have also been  used in previous work in the literature \citep{Martens13asynchronous}. The crossover and mutation chosen, SBX and polynomial, have been also used previously by other authors in \citep{Durillo08masterslave}.

@@

The termination criterion used is the execution time: 25 seconds for dimension 512 and 100 (four times more) for 2048. We have used the time instead the number of evaluations firstly because our hypothesis argue that the time saved in crossover in mutation can be spent on improving the sub-populations and more operations and migrations can be achieved. Also, we are using different number of islands (with different sub-population sizes) that could lead to different execution times, so it would be difficult to compare different times and quality solutions at the same time.

@@

The ECJ framework \citep{ECJ} has been used to run the
experiments. Specific operators have been developed as new modules for
ECJ, and they can be downloaded from our GitHub repository under a
LGPL V3 License
\footnote{\url{https://https://github.com/hpmoon/hpmoon-islands}}. The
island model has been executed synchronously, using the ECJ Internal
Island Model, in a CentOS 5.4 machine with Intel(R) Xeon(R) CPU E5520
@2.27GHz, 16 GB RAM, with Java Version 1.6.0\_16.

@@

%DESCRIPTION OF THE RESULTS
Results show that there is a difference in performance when using 512 or 2048 dimensions. Although in some cases the overlapped methods obtain quality metrics values better or significantly equal than the obtained by the baseline, it is clear that 512 dimensions is not a value high enough to apply these methods. This contradicts the results obtained in our previous work in a shared-memory mono-processor version \citep{Garcia16hpmoon}, where the overlapping method obtained better results than the baseline in both dimension lengths. This can be explained because the added migration latency between nodes requires more time than the time saved during the crossover and migration.

@@

%EXPLANATION OF THE RESULTS
This can be explained comparing the number non-dominated solutions in each island and the average number of generations (Table \ref{tab:sols512} and Table \ref{tab:sols2048}). With both chromosome lengths, increasing the number of islands implies more generations with all the methods (logically, as there are less individuals in each island). But also, the co-evolutionary methods approach to the number of generations with the baseline when increasing the number of islands. However, with $L=2048$ the migration time is enough big to almost not improve the number of generations with respect to the baseline, even in the lower number of islands, as it happens with $L=512$. Therefore, more generations do not necessarily improve the solution of the global PF, but focusing on different elements of the chromosome. As indicated previously, every island is unaware of the other islands PFs, and they are trying to optimize their solutions independently. With respect to the average number of solutions in each island, there is a clear difference in number with the baseline with $L=512$, where this value is in the most of the cases, less than a half. The number of non-dominated solutions also improves a better Spread indicator, where the baseline almost always obtains better values %TENGO QUE COMENTAR MAS LOS RESULTADOS Y ENTENDERLO TODO Y JUSTIFICARLO FINAMENTE

Finally, as happened in \citep{Garcia16hpmoon}, results show that all the quality indicators decrease when the number of islands is increased, as the  sizes of the subpopulations are smaller. This is consistent with the claim by Dorronsoro et al. \citep{Dorronsoro13superlinear}, who found that cooperative coevolutionary MOEAs work better on bigger populations (more than 100 individuals).

@@

\begin{acks}
Acknowledgements
\end{acks}

TIN2014-56494-C4-3-P y TIN2015-67020-P (Ministerio de Economía y Competitividad), PROY-PP2015-06 (Plan Propio 2015 UGR),  y MSTR (PRY142/14, Fundaci\'on P\'ublica Andaluza Centro de Estudios Andaluces en la IX Convocatoria de Proyectos de Investigaci\'on).

>>>>

\def\volumeyear{2016}

@@

%\author{Pablo Garc\'ia-S\'anchez, Julio Ortega, Jes\'us Gonz\'alez, \\Pedro Angel Castillo and Juan Juli\'an Merelo\affilnum{1}}
\author{ANONYMOUS FOR BLIND REVIEW \affilnum{1}}
\affiliation{\affilnum{1} ANONYMOUS FOR BLIND REVIEW}
\affiliation{\affilnum{1}Dept. of Computer Architecture and Computer Technology, University of Granada, Spain}

@@

%\corrauth{ANONYMOUS FOR BLIND REVIEW.}

@@

In our preliminary work in \citep{Garcia16hpmoonANONYMOUS} we tried to demonstrate that using overlapped sections of the chromosome in a coevolutionary multi-objective island-based algorithm can improve the quality of the solutions in the same computing time when the number of islands increases. We discovered that the most adequate overlapping size may vary depending on a relation between the number of islands and the individual size. Moreover, the experiments were conducted in a mono-processor and synchronous island model. This motivates us to continue this research line, but this time, proposing a new method to automatically adapt the size of the overlapped sections, but also, performing all the experiments in a real cluster with up to 128 nodes, with extra number of islands.

@@

The rest of the paper is organized as follows: after a background in parallelization in MOEAs, 
the compared algorithms and the used methodology are presented. %PABLO: no puedo usar \ref{sec:met} porque la plantilla de esta revista no lo coge :/ (y además no tienen número)
Then, the results of the experiments are shown, followed by conclusions and suggestions for future work lines.

@@

In our previous work \citep{Garcia16hpmoonANONYMOUS} we used some of the previous ideas to compare two co-evolutionary dMOEAs. The first one divided the chromosome in $P$ sections, being $P$ the number of islands. Every island $p$ only performed the mutation and crossover in that part ($p_{th}$) of the chromosome, while the fitness was calculated using the whole individual. After a certain number of generations, individuals were migrated randomly to other islands. The performance metrics were calculated at the end of the run. The second method, coevolution with overlapping islands, used the $p_{th-1}$ and $p_{th+1}$ sections of each chromosome, besides the $p_{th}$ part. During the same amount of time both methods obtained better results than a baseline algorithm that dealt with the whole chromosome in each island for crossover and mutation. We discovered that the performance using one or another method depends on the number of sections of the individuals and the number of islands used. This motivates us to find a new adaptive method to select this number of sections of the chromosome to use, depending on the number of islands. Moreover, previous experiments were performed on a single processor synchronous island model with a limited number of islands (8, 32 and 128). In this paper,  8, 16, 32, 64 and 128 islands have been used, and this time, the experiments have been run on a real cluster. Therefore, at the same time we are proposing an adaptive method to divide the individual search space, we are validating the previous methods.

@@

%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%  METHODOLOGY  %%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%

\section{Coevolutionary algorithms compared}
\label{sec:coevo}

In this paper we analyse several coevolutionary algorithms we have implemented by using a parallel multi-objective baseline algorithm (B) as a framework. Here, we have used the NSGA-II algorithm as the baseline algorithm. The population is distributed among $P$ islands. After a fixed number of generations, one individual in a given island is migrated to another random island, thus avoiding to synchronize the global Pareto Front (PF) every certain number of generations as other methods described in Section 2 do. At the end of the run, the Pareto Fronts of all islands are aggregated into a new one, and the quality measures are evaluated.
Different coevolutionary alternatives can be devised to evolve the subpopulations according to the decision space to be explored by each islands. Here we will consider three methods in which each island only performs crossover and mutation in specific sections of the individuals.

%This algorithm is a regular NSGA-II algorithm distributed along a number $P$ of islands. After a fixed number of generations, one individual is migrated to another random island. At the end of the run, PFs of all islands are aggregated in a new one in order to compute the quality measures. This baseline method has been chosen as it does not require synchronization of the global PF every certain number of generations, as other methods described in Section 2 do.
%% some justification on why this was chosen - JJ PABLO: Added 
%
%% And some liaison to next subsection - JJ
%This method will be compared to different coevolutionary approaches that are similar to the baseline, with the exception on the decision space to explore in each island.
%Three different coevolutionary methods have been used. In all these methods, each island only performs crossover and mutation in specific sections of the individuals.

As in the baseline, every certain number of generations, an individual is migrated to another random island.  In the new island, this individual will be considered as one of the others in the island, crossed and mutated in the same way, depending on the island identifier. Note that, on the contrary of other works such as the ones described by Talbi et al. \citep{Talbi08Parallel} all the islands deal with complete chromosomes for fitness calculation, so our approach can deal with no decomposable problems. The proposed methods  are as follows:

\subsection{Coevolution with disjoint islands (D)} 
In this approach, each individual of size $L$ is split into $P$ chunks of size $L/P$. Every island $p$ only performs crossover and mutation on the $p_{th}$ part of the individuals. Figure \ref{fig:disjoint} describes this approach.

\subsection{Coevolution with overlapping islands (O)}
This approach is similar to the previous one, but every island also uses the $p+1$ and $p-1$ (module size) chunks of the individual for crossover and migration. Therefore, some kind of overlapping of the crossed and mutated parts exists between islands. Figure \ref{fig:overlapping} shows the affected parts of the individuals in
each island.

\subsection{Adaptive coevolution with overlapping islands (A)} 
As in the previous method, the sections to address are overlapped, but instead using one extra section on each side of the $p$ section ($p+1$ and $p-1$), it uses $c$ fragments on each side ($p+c$ and $p-c$), being $c$ a value depending on the number of islands. As a first approach to automatically calculate this value the results shown in \citep{Garcia16hpmoonANONYMOUS} have been used as a base to obtain $c$. In that work, the overlapped method ($c=1$) obtained better results when the number of islands were $>$8. So, we have used the formula $c=round(0.2*P-1)/2$ to calculate the extra sections to overlap. Therefore, when $P=8$ then $c=0$ (equivalent to D), when $P=16$ then $c=1$ (equivalent to O), and so on. Figure \ref{fig:adaptive} explains this method.

This section describes the quality indicators used and the experimental setup. The chosen quality indicators, are:

@@

These metrics have been used extensively, especially in some of the papers presented in Section 2.

The previously described four approaches have been run with two different chromosome lengths ($L$): 512 and 2048. Different number of islands ($P$) have also been compared: 8, 16, 32, 64 and 128. This maximum number of islands have also been  used in previous work in the literature \citep{Martens13asynchronous}. The crossover and mutation chosen, SBX and polynomial, have also been  used previously by other authors in \citep{Durillo08masterslave}.

@@

Termination criterion used has been the running time: 25 seconds for dimension 512 and 100 (four times more) for 2048. We have used the time instead the number of evaluations firstly because our hypothesis argue that the time saved in crossover and mutation can be spent on improving the sub-populations and more operations and migrations can be achieved. Also, we are using different number of islands (with different sub-population sizes) that could lead to different execution times, so it would be difficult to compare different times and quality solutions at the same time.

@@

The ECJ framework \citep{ECJ} has been used to run the
experiments. Specific operators have been developed as new modules for
ECJ, and they can be downloaded from our GitHub repository under a
LGPL V3 License
%\footnote{\url{https://github.com/hpmoon/hpmoon-islands}}. 
\footnote{\url{https://ANONYMOUS-REPOSITORY}}. 
The
island model has been executed synchronously, using the ECJ Internal
Island Model, in a CentOS 5.4 machine with Intel(R) Xeon(R) CPU E5520
@2.27GHz, 16 GB RAM, with Java Version 1.6.0\_16.

@@

%DESCRIPTION OF THE RESULTS
Results show that there is a difference in performance when using 512 or 2048 dimensions. Although in some cases the overlapped methods obtain quality metrics values better or significantly equal than the obtained by the baseline, it is clear that 512 dimensions is not a value high enough to apply these methods. This contradicts the results obtained in our previous work in a shared-memory mono-processor version \citep{Garcia16hpmoonANONYMOUS}, where the overlapping method obtained better results than the baseline in both dimension lengths. This can be explained because the added migration latency between nodes requires more time than the time saved during the crossover and migration.

@@

%EXPLANATION OF THE RESULTS
This can be explained comparing the number of non-dominated solutions in each island and the average number of generations (Table \ref{tab:sols512} and Table \ref{tab:sols2048}). With both chromosome lengths, increasing the number of islands implies more generations with all the methods (logically, as there are less individuals in each island). But also, the co-evolutionary methods approach to the number of generations with the baseline when increasing the number of islands. However, with $L=2048$ the migration time is enough big to almost not improve the number of generations with respect to the baseline, even in the lower number of islands, as it happens with $L=512$. Therefore, more generations do not necessarily improve the solution of the global PF, but focusing on different elements of the chromosome. As indicated previously, every island is unaware of the other islands PFs, and they are trying to optimize their solutions independently. With respect to the average number of solutions in each island, there is a clear difference in number with the baseline with $L=512$, where this value is in the most of the cases, less than a half. The number of non-dominated solutions also improves a better Spread indicator, where the baseline almost always obtains better values %TENGO QUE COMENTAR MAS LOS RESULTADOS Y ENTENDERLO TODO Y JUSTIFICARLO FINAMENTE

Finally, as happened in \citep{Garcia16hpmoonANONYMOUS}, results show that all the quality indicators decrease when the number of islands is increased, as the  sizes of the subpopulations are smaller. This is consistent with the claim by Dorronsoro et al. \citep{Dorronsoro13superlinear}, who found that cooperative coevolutionary MOEAs work better on bigger populations (more than 100 individuals).

@@

\begin{acks}
ANONYMOUS
%This work has been partially funded by projects TIN2014-56494-C4-3-P y TIN2015-67020-P (Spanish Ministry of Economy and Competitivity), PROY-PP2015-06 (Plan Propio 2015 UGR),  and  MSTR (PRY142/14, Fundaci\'on P\'ublica Andaluza Centro de Estudios Andaluces en la IX Convocatoria de Proyectos de Investigaci\'on).
\end{acks}
